Associations between Chlorophyll *a* and various Microcystin-LR Health Advisory

### 2015-12-04 manuscript/manuscript.Rmd
closes #5 plus some changes to discussion and abstract from addition of thresholds table.

#Agreement Tables
child_conmat<-conmat_to_df(table(chla=nla_dat$CHLA>=mc_chla[1],
                                 mcsyt=nla_dat$MCYST_TL_UGL>=0.3))
who_drink_conmat<-conmat_to_df(table(nla_dat$CHLA>=mc_chla[2],
                        nla_dat$MCYST_TL_UGL>=1))
adult_conmat<-conmat_to_df(table(nla_dat$CHLA>=mc_chla[3],
                    nla_dat$MCYST_TL_UGL>=1.6))
who_rec_conmat<-conmat_to_df(table(nla_dat$CHLA>=mc_chla[4],
                      nla_dat$MCYST_TL_UGL>=2))
total_accuracy<-NULL
total_accuracy$epa_child<-round((child_conmat[1,1]+child_conmat[2,2])/
  sum(child_conmat),2)
total_accuracy$who_drink<-round((who_drink_conmat[1,1]+who_drink_conmat[2,2])/
  sum(who_drink_conmat),2)
total_accuracy$epa_adult<-round((adult_conmat[1,1]+adult_conmat[2,2])/
  sum(adult_conmat),2)
total_accuracy$who_rec<-round((who_rec_conmat[1,1]+who_rec_conmat[2,2])/
  sum(who_rec_conmat),2)

@@

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect against these impacts, several health advisory levels have been set for some toxins.  In particular, one of the more common toxins, microcystin, has several advisory levels set for drinking water and recreational use.  However, compared to other water quality measures, field measurements of microcystin are not commonly available due to cost and advanced understanding required to interpret results.  Addressing these issues will take time and resources.  Thus, there is utility in finding indicators of microcystin that are already widely available, can be estimated quickly and *in situ*, and used as a first defense against high levels of microcystin.  In particular, chlorophyll *a* is very commonly measured, can be estimated *in situ*, and has been shown to be positively associated with microcystin.  In this paper, we use this association to provide estimates of chlorophyll *a* concentrations that if exceeded would be indicative of a higher probability of exceeding select health advisory concentrations for microcystin-LR.  Using the 2007 National Lakes Assessment and a conditional probability approach, we identify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA health advisories for drinking water as well as the World Health Organization levels for drinking water and recreational use.  For the specific advisory microcystin concentrations of 0.3, 1, 1.6, and 2, we find chlorophyll *a* concentrations of `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]`, respectively.  When managing for these various microcystin levels, exceeding these reported chlorophyll *a* concentrations should be a trigger for further testing and possible management action.

@@

Over the last decade or so numerous events and legislative activities have raised the public awareness of harmful algal blooms [@jetoo2015toledo; @rinta2009lake; @HABHRCA2014], and in response the US Environmental Protection Agency (USEPA) has recently released suggested microcystin-LR (one of the more common toxins) concentrations that would trigger health advisories [@mcelhiney2005detection; @zurawell2005hepatotoxic; @usepa2015drinking].  Additionally, the World Health Organization (WHO) has had proposed advisory levels for drinking water and a range of recreational risk levels [@world2003cyanobacterial; @chorus1999toxic].   While these levels and associated advisories are likely to help mitigate the impacts from harmful algal blooms, they are not without complications.

One of these complications is that they rely on available measurements of microcystin-LR.  While laboratory testing remains the gold standard for quantifying microcystin-LR concentrations in water samples, several field test kits have been developed.  Even though field tests provide a much needed means for rapid assessment, they are not yet widely used and are moderately expensive (approximately $150-$200 depending on specific kit) with a limited shelf life (typically one year) [@james2011environmental; @aranda2015evaluation].  Additionally, each technique requires nuanced understanding of the detection method (e.g., limit of detection, specific microcystin variants being measured, and sampling protocol).

@@

##Data
We used the 2007 NLA water quality and microcystin-LR concentration data [@usepa2009national]. These data represent a snapshot of water quality from the summer of 2007 for the conterminous United States. Data on chlorophyll *a* and microcystin-LR concentrations are available for `r nrow(na.omit(nla_dat$CHLA,nla_dat$MCYST_TL_UGL))` lakes.

@@

Conditional probability analysis provides information about the probability of observing one event given another event has also occured.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisories changes as chlorophyll *a* increases in a lake.  We expect to find higher chlorohpyll *a* concentrations to be associated with higher probabilities of exceeding the microcystin-LR health advisory levels.  We also caclulated bootstrapped 95% confidence intervals (CI) using 1000 bootstrapped samples.  Thus, to identify chlorophyll *a* concentrations of concern we identfiy the value of the upper 95% CI across a range of conditional probabilities of exceeding each health advisory level.  As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.  A more detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for greater detail.

@@

#Discussion
The association between Log10 microcystin-LR and Log10 chlorophyll *a* shows a wedge pattern (Figure \ref{fig:chla_micro_scatter}).  This indicates that higher concentrations of microcystin-LR almost always co-occur with higher concentrations of chlorophyll *a* yet the inverse is not true.  Higher chlorophyll *a* is not necessarily predictive of higher microcystin-LR concentrations; however, chlorophyll *a* may be predictive of the probability of exceeding a certain concentration.

This is the case as the probability of exceeding each of the four tested health advisory levels increases as a function of chlorophyll *a* concentration (Figure \ref{fig:multi_cp_plot}).  We use this association to identify chlorophyll *a* concentrations that are associated with greater than even odds of exceeding a given health advisory level (Table \ref{tab:mc_chla_table}).  These represent `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[1]])/length(nla_dat$CHLA)),1)`%, `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[2]])/length(nla_dat$CHLA)),1)`%,
`r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[3]])/length(nla_dat$CHLA)),1)`%, and `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[4]])/length(nla_dat$CHLA)),1)`% of sample lakes for the USEPA Child, the WHO Drinking, the USEPA Adult, and the WHO recreational levels, respecitvely.

@@

A second potential use is to identify possible bloom events from historical data.  As harmful algal blooms are made up of many species and have various mechanisms responsible for adverse impacts (e.g. toxins, hypoxia, odors), there is no single definition of a bloom.  For cHABs one approach has been to identify an increase over a baseline concentration of phycocyanin [@miller2013spatiotemporal].  This is a useful approach for targeted studies, but phycocyanin is also not always available and measures the predominance of cyanobacterial pigments and not toxins. Using our chlorophyll *a* cutoffs provides a value that is more directly associated with microcystin-LR and can be used to classify lakes, from past surveys, as having bloomed.

Lastly, using chlorophyll *a* is not meant as a replacement for testing of microcystin-LR.  It should be used when other, direct measurements of cyanotoxins are not available.  In those cases, which are likely to be common at least in the near future, using a more ubiquitous measurement, such as chlorophyll *a* will provide a reasonable proxy for the probability of exceeding a microcystin-LR health advisory level and provide better protection against adverse effects in both drinking and recreational use cases.

>>>>

#Agreement Tables
child_conmat<-conmat_to_df(table(chla=nla_dat$CHLA>=mc_chla[5,1],
                                 mcsyt=nla_dat$MCYST_TL_UGL>=0.3))
who_drink_conmat<-conmat_to_df(table(nla_dat$CHLA>=mc_chla[5,2],
                        nla_dat$MCYST_TL_UGL>=1))
adult_conmat<-conmat_to_df(table(nla_dat$CHLA>=mc_chla[5,3],
                    nla_dat$MCYST_TL_UGL>=1.6))
who_rec_conmat<-conmat_to_df(table(nla_dat$CHLA>=mc_chla[5,4],
                      nla_dat$MCYST_TL_UGL>=2))
total_accuracy<-NULL
total_accuracy$epa_child<-round((child_conmat[1,1]+child_conmat[2,2])/
  sum(child_conmat),2)
total_accuracy$who_drink<-round((who_drink_conmat[1,1]+who_drink_conmat[2,2])/
  sum(who_drink_conmat),2)
total_accuracy$epa_adult<-round((adult_conmat[1,1]+adult_conmat[2,2])/
  sum(adult_conmat),2)
total_accuracy$who_rec<-round((who_rec_conmat[1,1]+who_rec_conmat[2,2])/
  sum(who_rec_conmat),2)

@@

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect against these impacts, several health advisory levels have been set for some toxins.  In particular, one of the more common toxins, microcystin, has several advisory levels set for drinking water and recreational use.  However, compared to other water quality measures, field measurements of microcystin are not commonly available due to cost and advanced understanding required to interpret results.  Addressing these issues will take time and resources.  Thus, there is utility in finding indicators of microcystin that are already widely available, can be estimated quickly and *in situ*, and used as a first defense against high levels of microcystin.  Chlorophyll *a* is commonly measured, can be estimated *in situ*, and has been shown to be positively associated with microcystin.  In this paper, we use this association to provide estimates of chlorophyll *a* concentrations that are indicative of a higher probability of exceeding select health advisory concentrations for microcystin-LR.  Using the 2007 National Lakes Assessment and a conditional probability approach, we identify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA health advisories for drinking water as well as the World Health Organization levels for drinking water and recreational use and identify a range of chlorophyll *a* thresholds.  A 50% chance of exceeding one of the specific advisory microcystin concentrations of 0.3, 1, 1.6, and 2 $\mu$g/L is associatied with chlorophyll *a* concentration thresholds of `r mc_chla[5,1]`, `r mc_chla[5,2]`, `r mc_chla[5,3]`, and `r mc_chla[5,4]`, respectively.  When managing for these various microcystin levels, exceeding these reported chlorophyll *a* concentrations should be a trigger for further testing and possible management action.

@@

Over the last decade, numerous events and legislative activities have raised the public awareness of harmful algal blooms [@jetoo2015toledo; @rinta2009lake; @HABHRCA2014]. In response the US Environmental Protection Agency (USEPA) has recently released suggested microcystin-LR (one of the more common toxins) concentrations that would trigger health advisories [@mcelhiney2005detection; @zurawell2005hepatotoxic; @usepa2015drinking].  Additionally, the World Health Organization (WHO) has proposed microcystin advisory levels for drinking water and a range of recreational risk levels [@world2003cyanobacterial; @chorus1999toxic].   While these levels and associated advisories are likely to help mitigate the impacts from harmful algal blooms, they are not without complications.

One of these complications is that they rely on available measurements of microcystin-LR.  While laboratory testing (e.g., chromatography) remains the gold standard for quantifying microcystin-LR concentrations in water samples, several field test kits have been developed.  Even though field tests provide a much needed means for rapid assessment, they are not yet widely used and are moderately expensive (approximately $150-$200 depending on specific kit) with a limited shelf life (typically one year) [@james2011environmental; @aranda2015evaluation].  Additionally, each technique requires nuanced understanding of the detection method (e.g., limit of detection, specific microcystin variants being measured, and sampling protocol).

@@

##Data
We used the 2007 NLA chlorophyll *a* and microcystin-LR concentration data [@usepa2009national]. These data represent a snapshot of water quality from the summer of 2007 for the conterminous United States and were collected as part of an ongoing probabilistic monitoring program [@usepa2009national]. Data on chlorophyll *a* and microcystin-LR concentrations are available for `r nrow(na.omit(nla_dat$CHLA,nla_dat$MCYST_TL_UGL))` lakes.

@@

Conditional probability analysis provides information about the probability of observing one event given another event has also occured.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisories changes as chlorophyll *a* increases in a lake.  We expect to find higher chlorohpyll *a* concentrations to be associated with higher probabilities of exceeding the microcystin-LR health advisory levels.  We also caclulated bootstrapped 95% confidence intervals (CI) using 1000 bootstrapped samples.  Thus, to identify chlorophyll *a* concentrations of concern we identfiy the value of the upper 95% CI across a range of conditional probabilities of exceeding each health advisory level.  As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.  A more detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for greater detail.  Lastly, all analyses were conducted using R version 3.2.2 and code and data from this analysis are freely available as an R package at [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

@@

#Discussion
The association between Log10 microcystin-LR and Log10 chlorophyll *a* shows a wedge pattern (Figure \ref{fig:chla_micro_scatter}).  This indicates that, in general, higher concentrations of microcystin-LR almost always co-occur with higher concentrations of chlorophyll *a* yet the inverse is not true.  Higher chlorophyll *a* is not necessarily predictive of higher microcystin-LR concentrations; however, chlorophyll *a* may be predictive of the probability of exceeding a certain threshold.

This is the case as the probability of exceeding each of the four tested health advisory levels increases as a function of chlorophyll *a* concentration (Figure \ref{fig:multi_cp_plot}).  We use this association to identify chlorophyll *a* concentrations that are associated with a range of probabilities of exceeding a given health advisory level (Table \ref{tab:mc_chla_table}).  For the purposes of this discussion we focus on a conditional probability of 50% or greater (i.e. greater than even odds to exceed a health advisory level). The 50% conditional probability chlorophyll *a* thresholds represents `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[5,1]])/length(nla_dat$CHLA)),1)`%, `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[5,2]])/length(nla_dat$CHLA)),1)`%,
`r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[5,3]])/length(nla_dat$CHLA)),1)`%, and `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[5,4]])/length(nla_dat$CHLA)),1)`% of sample lakes for the USEPA Child, the WHO Drinking, the USEPA Adult, and the WHO recreational levels, respecitvely.

@@

A second potential use is to identify past bloom events from historical data.  As harmful algal blooms are made up of many species and have various mechanisms responsible for adverse impacts (e.g. toxins, hypoxia, odors), there is no single definition of a bloom.  For cHABs one approach has been to identify an increase over a baseline concentration of phycocyanin [@miller2013spatiotemporal].  This is a useful approach for targeted studies, but phycocyanin is also not always available and measures the predominance of cyanobacterial pigments and not toxins. Using our chlorophyll *a* cutoffs provides a value that is more directly associated with microcystin-LR and can be used to classify lakes, from past surveys, as having bloomed.

Lastly, using chlorophyll *a* is not meant as a replacement for testing of microcystin-LR or other toxins.  It should be used when other, direct measurements of cyanotoxins are not available.  In those cases, which are likely to be common at least in the near future, using a more ubiquitous measurement, such as chlorophyll *a* will provide a reasonable proxy for the probability of exceeding a microcystin-LR health advisory level and provide better protection against adverse effects in both drinking and recreational use cases.
